---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# move2imu

<!-- badges: start -->
[![R-CMD-check](https://github.com/robe2037/move2imu/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/robe2037/move2imu/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/robe2037/move2imu/graph/badge.svg)](https://app.codecov.io/gh/robe2037/move2imu)
[![CRAN status](https://www.r-pkg.org/badges/version/move2imu)](https://CRAN.R-project.org/package=move2imu)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

## Design

* Keep frequency and axes per element or for the vector? -> per element as it changes within study
* Allow NA's? -> yes in vector , not in burst?
* Are start time part of vector? -> probably not as move2 already keeps it
* Calibrated (in G) or Uncalibrated -> might be covered by storing units
* We assume constant sampling freq in a burst


## Installation

You can install the development version of move2imu like so:

``` r
# install.packages("remotes")
remotes::install_github("robe2037/move2imu")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(move2imu)
## basic example code
```

