---
title: "Calibrating acceleration bursts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating acceleration bursts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "80%"
)

local({
  hook_output <- knitr::knit_hooks$get("output")
  knitr::knit_hooks$set(output = function(x, options) {
    n <- options$out.lines
    if (!is.null(n)) {
      lines <- strsplit(x, "\n", fixed = TRUE)[[1]]
      if (length(lines) > n) {
        x <- paste(c(lines[seq_len(n)], "#> ..."), collapse = "\n")
      }
    }
    hook_output(x, options)
  })
})

Sys.setlocale("LC_TIME", "en_US.utf8")
```

Many accelerometer tags store their measurements as raw integers produced by
the analog-to-digital converter (ADC). By themselves, these values have no
physical units. To convert the raw ADC values to interpretable acceleration
units (e.g., m/s²), we need to define the transformation that maps each
ADC value to its corresponding value in the target unit.

For acceleration, this is a simple linear transformation, requiring:

1. An **offset**, which is the ADC value that corresponds to 0 acceleration;
2. A **slope**, which defines the output unit increment that corresponds to
  a one-unit ADC increment.

Further, different tag manufacturers and generations may define the axis
space differently. That is, one tag manufacturer may define the y-axis to
point in the opposite direction of another manufacturer. To account for this,
a transformation function also takes an **orientation** indicator, which 
identifies the direction that each axis points. This allows you to ensure
that all tags being considered together use the same axis orientation
definitions.

For a set of acceleration bursts that all come from the same tag manufacturer
and tag generation, the transformation from ADC to physical units is likely
to be essentially identical. However, for a more complex dataset with 
long-running deployments and/or multiple tag manufacturers, each tag will
need its own transformation.

move2imu facilitates this transformation process with two functions:

- `acc_calibration()` specifies the parameters for a set of transformation
functions
- `transform_imu()` applies that specification to an `acc` vector,
yielding a vector with proper physical units.

We'll work through this with the example `albatrosses()` dataset,
which contains raw data collected with e-obs tags.

```{r load, message=FALSE}
library(move2imu)
library(move2)
library(dplyr)
```

First, we'll load the data and extract the acceleration burst data. 
This sample dataset is shipped with move2imu:

```{r raw}
alb <- albatrosses()

alb <- alb |>
  mutate(acceleration = as_acc(alb))

head(alb$acceleration)
```

From the summary, we can see that no units have yet been attached to the
values:

```{r}
summary(alb$acceleration)
```

Note as well that the raw acc values are in the thousands, which would be 
unreasonable were these values already in m/s² or *g*.

We can confirm with `imu_units()`:

```{r raw_units}
imu_units(alb$acceleration)
```

Before we can truly understand these data, we'll need to convert
them to physical units.

## Tag manufacturer calibration

The easiest way to convert ADC values is with the tag manufacturer
default transformation settings.

move2imu currently has built-in support for two tag manufacturers:
[e-obs](https://e-obs.de/) and [Ornitela](https://www.ornitela.com/).

These sample data were collected with e-obs tags.
The default calibration suggested by e-obs differs by tag generation. Thus,
to build a calibration, we need to specify both the tag manufacturer and the 
tag ID, which identifies the tag generation. In this case, imagine that these
data were collected with the latest generation of e-obs tags (tags with IDs
greater than 4117):

```{r single}
cal <- acc_calibration(manufacturer = "eobs", tag_id = 4266)
cal
```

Then, we simply use `transform_imu()` to apply this transformation to
the albatross acceleration bursts:

```{r}
alb$acc_cal <- transform_imu(alb$acceleration, calibration = cal)

head(alb$acc_cal)
```

By default, `acc_calibration()` converts to m/s². If you
instead want your data in *g*, you can convert with `set_imu_units()`:

```{r}
set_imu_units(alb$acc_cal, "standard_free_fall")
```

Alternatively, we could have also specified the units in the calibration itself:

```{r}
acc_calibration(
  manufacturer = "eobs",
  tag_id = 4266,
  units = "standard_free_fall"
)
```

Note the difference between these two approaches: `set_imu_units()` 
converts *between* compatible units (e.g.,
m/s² to *g*) or *assigns* new units, whereas `transform_imu()`
*transforms* raw values to their physical counterpart. 

Thus, if you have raw
values in your data, you must first use `transform_imu()` to convert the raw
ADC values to meaningful units using the provided calibration. Otherwise,
you will end up simply assigning units to raw values without changing the 
values themselves.

```{r}
alb$acceleration[2]

set_imu_units(alb$acceleration[2], "standard_free_fall")
```

## Vectorized calibration

The problem with the above approach is that it assumes that *all* tags used
in the study had ID `4266` (generation 3). If all the albatrosses tagged in the 
study used generation 3 e-obs tags, then this wouldn't be a problem.

However, e-obs has revised the default calibration over time, so earlier 
generations of tags don't necessarily have the same transformation function.
(You can see the default calibration specifications for e-obs tags with 
`eobs_default_specs()`.)

In these cases, you can build a vector of per-burst calibration specifications,
which when passed to `transform_imu()` will be matched to bursts by index.

For instance, imagine that the e-obs tag IDs were stored in the original 
`move2` object like so:

```{r}
# Example IDs that span e-obs generations
fake_ids <- c(1000, 2000, 2500, 2501, 2502, 4200, 4201, 4202, 4203)

# Set example IDs in albatross track data
alb <- mt_set_track_data(
  alb,
  mutate(mt_track_data(alb), tag_id = fake_ids)
)

mt_track_data(alb)
```

We could use this track-level metadata to build a vectorized set of calibration
functions based on the tag ID:

```{r}
cals <- acc_calibration("eobs", tag_id = mt_track_data(alb)$tag_id)

# Reattach calibrations to each track
alb <- mt_set_track_data(
  alb,
  mutate(mt_track_data(alb), cal = cals)
)

mt_track_data(alb)$cal
```

The only issue is that these calibrations are linked to each track, but our
acceleration bursts are stored as individual events in the *event* data of our
`move2` object.

To expand each calibration to its corresponding bursts (linking by track ID), 
we can use move2's `mt_as_event_attribute()`:

```{r, out.lines=10}
mt_as_event_attribute(alb, cal)$cal
```

Passing this to `transform_imu()` maps each burst to its corresponding
calibration function by index:

```{r}
transform_imu(alb$acceleration, mt_as_event_attribute(alb, cal)$cal)
```

In general, expanding within the call to `transform_imu()` is ideal, as the 
calibrations that are duplicated across bursts do not need to persist in 
memory afterwards.

## Custom calibration

For tags that come from manufacturers without built-in move2imu support, you can
build your own calibration function by manually specifying the offset and slope
for the transformation function:

```{r}
acc_calibration(offset = 2048, slope = 0.001)
```

If the calibration differs by axis, you can specify per-axis parameters:

```{r}
acc_calibration(offset_x = 2048, offset_y = 2040, slope = 0.001, orientation_y = -1)
```

You can also specify individual arguments to override a manufacturer 
default---for instance, if you have detailed information about some
calibration parameters but not others.

This can also be useful to standardize different tags that have defined their 
axes in different directions. For instance, you may want to redefine the y-axis 
orientation for Ornitela tags to match other tags used in the study if those
tags define the y-axis in the opposite direction:

```{r override}
cal1 <- acc_calibration("ornitela")
cal2 <- acc_calibration("ornitela", orientation_y = -1)

transform_imu(alb$acceleration[2], cal1)

transform_imu(alb$acceleration[2], cal2)
```

User-supplied values always take precedence over the manufacturer
defaults.

## External calibration specification

For studies with many tags, manually specifying detailed per-axis offset, slope,
and orientation values quickly becomes tedious. Instead, these
values can be generated once by a calibration procedure and
saved to an external file for later reuse.

move2imu supports the conversion of a `data.frame` that contains these calibration
parameters to an `acc_calibration` vector on a per-row basis 
with `as_acc_calibration()`.
Any columns in the `data.frame` that
correspond to the available arguments to `acc_calibration()` will be used
to build calibration functions.

For instance, for the following set of calibration specifications:

```{r cal_table_setup, include = FALSE}
cal_table <- tibble::tribble(
  ~tag_id, ~offset_x, ~offset_y, ~offset_z, ~slope_x, ~slope_y, ~slope_z, ~orientation_y,
  1000, 2049, 2047, 2050, 0.00271, 0.00269, 0.00270, -1,
  2000, 2048, 2050, 2046, 0.00270, 0.00271, 0.00269, -1,
  2500, 2050, 2047, 2052, 0.00221, 0.00220, 0.00219, 1,
  2501, 2047, 2049, 2048, 0.00220, 0.00219, 0.00220, 1,
  2502, 2049, 2046, 2051, 0.00219, 0.00221, 0.00220, 1,
  4200, 2048, 2049, 2047, 0.00195, 0.00196, 0.00196, 1,
  4201, 2050, 2047, 2049, 0.00196, 0.00195, 0.00195, 1,
  4202, 2046, 2051, 2048, 0.00196, 0.00196, 0.00197, 1,
  4203, 2049, 2048, 2050, 0.00195, 0.00197, 0.00196, 1
)
```

```{r cal_table_read, eval = FALSE}
cal_table <- read.csv("calibration.csv")
```

```{r cal_table_print}
cal_table
```

We can generate a vector of `acc_calibration` objects linked to their
corresponding `tag_id` like so:

```{r}
cals_df <- cal_table |>
  mutate(cals = as_acc_calibration(cal_table)) |>
  select(tag_id, cals)

cals_df
```

Then, we can join these calibrations onto our `move2` track data, matching by
tag ID:

```{r}
alb <- mt_set_track_data(
  alb,
  left_join(mt_track_data(alb), cals_df, by = "tag_id")
)
```

And transform by expanding our track-level calibrations to each of the
corresponding bursts in the event data:

```{r}
transform_imu(alb$acceleration, mt_as_event_attribute(alb, cals)$cals)
```

Now, each burst has been transformed specifically with the detailed calibration
specifications for the tag that produced it.
