---
title: "Computing acceleration metrics: An example workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing acceleration metrics: An example workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "80%"
)
```

After extracting IMU data into a vector of bursts (see
[Loading IMU data](move2imu.html#loading-imu-data)), we will typically want
to summarize the individual burst samples into a more interpretable
measure of activity. This vignette shows how to extract simple activity
metrics from `acc` bursts and reattach these metrics to corresponding location
data collected with GPS.

```{r load, message = FALSE}
library(move2imu)
library(move2)
library(dplyr)
library(ggplot2)
```

First, we'll load the albatross example data and extract the acceleration
bursts:

```{r alb}
alb <- albatrosses()
alb$acceleration <- as_acc(alb)
```

These data contain 12-second bursts recorded at 5 Hz across two axes.

```{r}
alb$acceleration
```

## Computing dynamic body acceleration

Dynamic body acceleration is a commonly used measure of
animal activity and is composed of two acceleration components:

- A *static* component, which stems from gravity or other sources of
constant acceleration (and therefore reflects the animal's posture).
- A *dynamic* component, which stems from the animal's own movements.

The dynamic component is a proxy for activity and, in many species, for energy
expenditure (Wilson et al., 2020). To estimate it, we must isolate and remove
the static acceleration to the best of our ability.
We'll demonstrate how to compute vectorial dynamic body acceleration (VeDBA)
on our albatross data.

First, we calibrate our data to units of *g* to make our DBA results more
interpretable (see the [calibration vignette](calibration.html) for details).
For this purpose, we suppose that all the birds in the study
carried first-generation e-obs tags set to low sensitivity.

```{r}
alb$acceleration <- transform_imu(
  alb$acceleration,
  acc_calibration(
    "eobs",
    tag_id = 1000,
    sensitivity = "low",
    units = "standard_free_fall"
  )
)
```

### Choosing a smoothing window

To separate the two acceleration components for a burst of data, we compute a
running mean over all the samples within a given window of time. This helps
to separate cyclical patterns (e.g. wingbeats) from stable acceleration signals.

In general, the window length should span several cycles of the primary
movement of interest, but remain short enough to avoid misclassifying
changes in posture as movements.

The figure below shows how the window length affects the decomposition of
static and dynamic acceleration. We've generated some example 20 Hz data on a
single acceleration axis showing an acceleration pattern for a hypothetical bird.
The bird shows a cyclical flapping pattern, glides, and then adjusts its
posture (e.g. to start a descent), producing an overall change in the static
gravitational acceleration vector on this axis.

On the left, the blue line shows the static acceleration estimated for the
given window size, shown as a shaded blue rectangle. When the window is shorter
than the flapping frequency, the running mean reproduces much of that
oscillation, thus incorporating it into the static component. This means 
we underestimate the portion of the signal that comes from dynamic sources.

As the window size increases (the middle panel), we average several oscillations
together, canceling them out. This leaves most of the oscillatory behavior
correctly in the dynamic component, producing a good estimate of the dynamic
acceleration. When the posture changes, we incorporate most of this signal
correctly into the static signal, because the frequency of the change is slow
relative to the running mean window.

When the window is too long, though, we begin to incorporate the postural
change across a long period of samples, thus treating some of the static
acceleration as movement.

Finally, the bottom row uses the mean of the entire recording as the static
estimate. This approach assumes that little to no postural change occurred
during the recording. As you can see, while this succesfully captures the
wingbeats in the dynamic component, it also classifies the postural change
as dynamic as well, overestimating the dynamic acceleration.

```{r window-demo, echo = FALSE, fig.width = 8, fig.height = 7, out.width = "100%"}
hz <- 20
secs <- seq(0, 20 - 1 / hz, by = 1 / hz)

# Posture: body pitch drops from 5 to -20 degrees over 1.5 s around 14 s. The
# static acceleration on the axis is the share of gravity at that pitch.
pitch_change <- pmin(pmax((secs - 14) / 1.5 + 0.5, 0), 1)
pitch <- 5 - 25 * pitch_change
true_static <- sin(pitch * pi / 180)

# Movement: 8 wingbeats with a 1 s period, from 2 to 10 s
flapping <- secs >= 2 & secs < 10
true_dynamic <- ifelse(flapping, 0.3 * sin(2 * pi * secs), 0)

set.seed(1)
z <- true_static + true_dynamic + rnorm(length(secs), sd = 0.01)

sim <- acc(
  list(cbind(X = 0, Y = 0, Z = z)),
  frequency = units::set_units(hz, "Hz"),
  start = as.POSIXct("2024-01-01", tz = "UTC")
)

window_labels <- c(
  "0.5" = "0.5 s (too short)",
  "2" = "2 s (about right)",
  "8" = "8 s (too long)",
  "burst" = "Burst mean"
)

# Window values as passed to static_acc(): seconds, or "burst"
as_window <- function(w) if (w == "burst") "burst" else as.numeric(w)

# Static and dynamic components of the Z axis for each window
components <- purrr::map(
  names(window_labels),
  function(w) {
    s <- bursts(static_acc(sim, window = as_window(w)))[[1]][, "Z"]
    d <- bursts(dynamic_acc(sim, window = as_window(w)))[[1]][, "Z"]

    bind_rows(
      data.frame(
        panel = "Static (posture)", t = secs, raw = z,
        estimate = s, truth = true_static
      ),
      data.frame(
        panel = "Dynamic (movement)", t = secs, raw = NA,
        estimate = d, truth = true_dynamic
      )
    ) |>
      mutate(window = window_labels[[w]])
  }
) |>
  bind_rows() |>
  mutate(
    panel = factor(panel, c("Static (posture)", "Dynamic (movement)")),
    window = factor(window, window_labels)
  )

# Estimated and true VeDBA (mean absolute dynamic acceleration on one axis) for
# each window, over the samples that have an estimate
vedba_labels <- components |>
  filter(panel == "Dynamic (movement)", !is.na(estimate)) |>
  group_by(window, panel) |>
  summarize(
    vedba = mean(abs(estimate)),
    true_vedba = mean(abs(truth)),
    .groups = "drop"
  )

# One example window per row, centered on the peak of a wingbeat. The burst
# mean uses the whole burst for every sample.
example_t <- 6.25
half_widths <- suppressWarnings(as.numeric(names(window_labels))) / 2
example_windows <- data.frame(
  window = factor(window_labels, window_labels),
  panel = factor("Static (posture)", levels(components$panel)),
  xmin = ifelse(is.na(half_widths), min(secs), example_t - half_widths),
  xmax = ifelse(is.na(half_widths), max(secs), example_t + half_widths)
)
example_points <- components |>
  filter(panel == "Static (posture)", t == example_t)

ggplot(components, aes(x = t)) +
  geom_rect(
    data = example_windows,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE, fill = "#2a6fb0", alpha = 0.12
  ) +
  geom_line(aes(y = raw), colour = "grey75", linewidth = 0.3, na.rm = TRUE) +
  geom_line(aes(y = truth), linetype = "22", linewidth = 0.4) +
  geom_line(aes(y = estimate, colour = panel), linewidth = 0.7, na.rm = TRUE) +
  geom_point(
    data = example_points, aes(y = estimate),
    colour = "#2a6fb0", size = 2.5
  ) +
  geom_text(
    data = vedba_labels,
    aes(
      x = 20, y = Inf,
      label = sprintf("VeDBA %.2f g (true %.2f g)", vedba, true_vedba)
    ),
    hjust = 1, vjust = 1.4, size = 3, colour = "grey20"
  ) +
  scale_colour_manual(values = c("#2a6fb0", "#c4561d"), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
  facet_grid(window ~ panel, scales = "free_y") +
  labs(
    x = "Time (s)",
    y = "Acceleration (g)",
    subtitle = paste(
      "Dashed lines show the true values.",
      "Shaded bands show the window used to estimate the marked point."
    )
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

Thus, our ability to correctly estimate activity metrics, like VeDBA,
depend on an appropriate selection of the decomposition window.

Shepard et al. (2008) suggest at least a
3-second window for species whose stroke cycles are shorter than this.
Albatrosses mostly soar rather than flap, but the repeating movements that do
appear in these data (likely including occasional wingbeats) last about a
second or less, so we will adopt this guidance here.

To compute VeDBA, we simply use the `vedba()` function with the chosen
window.

```{r}
alb$vedba <- vedba(alb$acceleration, window = units::set_units(3, "s"))

alb$vedba
```

It's also possible to identify an appropriate window by computing VeDBA at
multiple window lengths and finding the point at which the curve flattens:

```{r sensitivity}
alb_acc <- alb$acceleration[!is.na(alb$acceleration)]

# Windows of 3 to 31 samples at 5 Hz
windows <- c(3, 5, 7, 11, 15, 21, 31) / 5

# Median VeDBA at each window size
v <- purrr::map_dbl(
  windows,
  function(w) median(as.numeric(vedba(alb_acc, window = w)))
)

d <- data.frame(window = windows, vedba = v)

ggplot(d, aes(x = window, y = vedba)) +
  geom_line() +
  geom_point() +
  labs(x = "Window (s)", y = "Median VeDBA (g)")
```

VeDBA rises steeply for windows up to about 1.5 seconds and changes little
beyond about 3 seconds, which supports choosing a 3-second window.

### Caveats

Some data are collected in shorter bursts that may not fit a meaningful window.
In these cases, you can pass `window = "burst"`, which estimates the static
component as the mean of each whole burst. However, these DBA results should
be interpreted with caution.

Furthermore, VeDBA is sensitive not only to the window length but also to
other features of the data, like sampling rate and number of axes.

These data were sampled at 5 Hz, which is quite slow relative to
most flapping frequencies. This limits the resolution with which we can
identify flapping flight, leading to underestimated VeDBA. Additionally, these
data were recorded on only 2 axes, while DBA metrics are often calculated on
data with 3. These VeDBA values thus shouldn't be compared to
measurements computed on 3-axis data.

The code here is intended to serve as a demonstration of a workflow for
computing these metrics rather than as a comprehensive analysis.

## Linking to locations

Often, the recording times of acceleration and GPS data don't match exactly.
In the albatross data, for instance, some GPS fixes are recorded up to a second
after the start of an acceleration burst. However, we may be interested in
associating acceleration metrics with nearby GPS locations.

To do so, we'll separate the VeDBA values
calculated above from the GPS fixes by removing rows with missing VeDBA values:

```{r}
alb_v <- alb |>
  filter(!is.na(vedba)) |>
  sf::st_drop_geometry() |>
  as_tibble() |>
  select(individual_local_identifier, timestamp, vedba) |>
  rename(vedba_time = timestamp)

head(alb_v)
```

Then, we'll isolate the GPS location rows by removing rows with empty geometries
and dropping the acceleration-based columns:

```{r}
alb_gps <- alb |>
  filter(!sf::st_is_empty(alb)) |>
  select(-acceleration, -vedba)
```

We can then join our VeDBA values to our GPS fix locations by each animal's
ID and recording timestamp. However, if we
require timestamps to match exactly, only 2 of the 9 fixes would have a match:

```{r}
left_join(
  alb_gps,
  alb_v,
  by = join_by(individual_local_identifier, timestamp == vedba_time)
) |>
  pull(vedba)
```

Instead, we can use a *rolling join* to match all records by a provided
inequality. Here, `closest(timestamp >= vedba_time)` links records by matching
each GPS fix timestamp to the most recent VeDBA record that started at or
before it, within the same individual:

```{r}
alb_joined <- left_join(
  alb_gps,
  alb_v,
  by = join_by(individual_local_identifier, closest(timestamp >= vedba_time))
)
```

A rolling join always takes the closest match, no matter how far away in time
it is. Before using the result, you may want to discard matches where the records
are far from each other in time. For instance, here we compute the time difference
for each match and keep only cases where the difference is at most 2 seconds:

```{r}
alb_joined <- alb_joined |>
  mutate(
    gap = as.numeric(timestamp - vedba_time, units = "secs"),
    vedba = if_else(gap <= 2, vedba, NA)
  )

alb_joined |>
  as_tibble() |>
  select(individual_local_identifier, timestamp, vedba_time, gap, vedba)
```

Here every fix is within a second of its burst, so all matches are kept.
The check matters more when records are sparse.

### Interpolating locations

These albatross data contain only one GPS fix per individual but five
acceleration bursts, so by joining to our GPS locations we end up discarding
some of the collected acceleration data. If we instead want to associate a
location with each *burst*, an alternative is to interpolate locations to the
burst times with `move2::mt_interpolate()`.

## Continuous data

Some tags record acceleration continuously, producing very long "bursts" in the
resulting `acc` vector. In these cases, `vedba()` would return a single value
for hours of data. Instead, you can separate the dynamic component of the 
whole recording with `dynamic_acc()`, then split it into intervals with
`split_imu()` and calculate VeDBA for each interval (see the examples in
`?dynamic_acc`). Each interval's start time can then be used to link it to
GPS fixes with a rolling join, as above.

Separating the dynamic component
before splitting means that only the samples at the two ends of the full
recording are lost when calculating the static acceleration, rather than the
samples at the edges of every interval.

## References

Shepard, E. L. C., Wilson, R. P., Halsey, L. G., Quintana, F., Gómez Laich,
A., Gleiss, A. C., Liebsch, N., Myers, A. E., & Norman, B. (2008).
Derivation of body motion via appropriate smoothing of acceleration data.
*Aquatic Biology*, 4, 235–241. <https://doi.org/10.3354/ab00104>

Wilson, R. P., Börger, L., Holton, M. D., Scantlebury, D. M., Gómez-Laich,
A., Quintana, F., et al. (2020). Estimates for energy expenditure in
free-living animals using acceleration proxies: A reappraisal. *Journal of
Animal Ecology*, 89, 161–172. <https://doi.org/10.1111/1365-2656.13040>
