---
title: "Programming with acceleration bursts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with acceleration bursts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

```{css, echo = FALSE}
.html-widget {
  margin-left: auto !important;
  margin-right: auto !important;
  width: 80% !important;
}
```

After extracting IMU data into a vector of bursts, we typically will want
to summarize the individual burst samples into a more interpretable
measure of activity. This vignette shows how to extract simple activity
metrics from `acc` bursts and reattach these metrics to corresponding location
data collected with GPS.

```{r load, message = FALSE}
library(move2imu)
library(move2)
library(dplyr)
library(units)
library(ggplot2)
```

First, we'll load the albatross example data and extract the acceleration
bursts:

```{r alb}
alb <- albatrosses()
alb$acceleration <- as_acc(alb)
```

These data contain 12 second bursts recorded at 5 Hz across two axes.

```{r}
alb$acceleration
```

## Computing Dynamic Body Acceleration

Dynamic body acceleration is a commonly used measure of
animal activity and is composed of two acceleration components:

- A *static* component, which stems from gravitation or other sources of 
constant acceleration (and therefore reflects the animal's posture).
- A *dynamic* component, which stems from the animal's own movements.

To estimate an animal's movement—a proxy for energy expenditure—we must
therefore isolate and remove the static acceleration to the best of our ability.
We'll demonstrate how to compute vectorial dynamic body acceleration (VeDBA)
on our albatross data.

First, we calibrate our data to units of *g* to make our DBA results more
interpretable. For this purpose, we imagine that all the birds in the study
were affixed with generation 3 e-obs tags.

```{r}
alb$acceleration <- transform_imu(
  alb$acceleration,
  acc_calibration("eobs", tag_id = 4266, units = "standard_free_fall")
)
```

### Choosing a smoothing window

To separate the two acceleration components for a burst of data, we compute a
running mean over all the samples within a given window of time. This helps
to separate cyclical patterns (e.g. wingbeats) from stable acceleration signals.

In general, the window length should span several cycles of the primary
movement of interest, but remain short enough to to avoid misclassifying
changes in posture as movements. Shepard et al. (2008) suggest at least a
3 second window for species with stroke movements faster than this value. Our
albatross wingbeats are likely to be must shorter than this, so we will
adopt this guidance here.

To compute VeDBA, we simply use the `vedba()` function with the determined
window.

```{r}
alb$vedba <- vedba(alb$acceleration, window = units::set_units(3, "s"))

alb$vedba
```

It's also possible to identify an appropriate window by computing VeDBA at
multiple window lengths and identifying the point at which the computed VeDBA
curve flattens.

```{r sensitivity}
acc <- alb$acceleration[!is.na(alb$acceleration)]

# Windows of 3, 5, 7, ... samples at 5 Hz
windows <- c(3, 5, 7, 11, 15, 21, 31) / 5

v <- purrr::map_dbl(
  windows,
  function(w) median(as.numeric(vedba(acc, window = w)))
)

d <- data.frame(window = windows, vedba = v)

ggplot(d, aes(window, vedba)) +
  geom_line() +
  geom_point() +
  labs(x = "Window (s)", y = "Median VeDBA (g)")
```

VeDBA rises steeply for windows up to about 1.5 seconds and changes little
beyond about 3 seconds, so this corroborates the selection of a 3 second window.

## Linking to locations

Often, the recording times of acceleration and GPS data don't match exactly.
In the albatross data, for instance, some GPS fixes are recorded up to a second
after the start of an acceleration burst. However, we may be interested in
associated acceleration metrics to nearby GPS locations.

To do so, we'll separate the VeDBA values
calculated above from the GPS fixes by removing rows with missing VeDBA values:

```{r}
alb_v <- alb |>
  filter(!is.na(vedba)) |>
  sf::st_drop_geometry() |>
  as_tibble() |>
  select(individual_local_identifier, timestamp, vedba) |>
  rename(v_time = timestamp)

head(alb_v)
```

Then, we'll isolate the GPS location rows by removing rows with empty geometries
and the acceleration-based columns:

```{r}
alb_gps <- alb |> 
  filter(!sf::st_is_empty(alb)) |> 
  select(-acceleration, -vedba)
```

We can then join our VeDBA values to our GPS fix locations by each animal's
ID and recording timestamp. However, if we
require timestamps to match exactly, only 2 of the 9 fixes would have a match:

```{r}
left_join(
  alb_gps,
  alb_v,
  by = join_by(individual_local_identifier, timestamp == v_time)
) |>
  pull(vedba)
```

Instead, we can use a *rolling join* to match all records by a provided
inequality. Here, `closest(timestamp >= v_time)` links records by matching
each GPS fix timestamp to the most recent VeDBA record that started at or
before it, within the same individual:

```{r}
alb_joined <- left_join(
  alb_gps,
  alb_v,
  by = join_by(individual_local_identifier, closest(timestamp >= v_time))
)
```

A rolling join always takes the closest match, no matter how far away in time
it is. Before using the result, check the time between each fix and its
burst, and discard matches that are too far apart for your purposes:

```{r}
alb_joined <- alb_joined |>
  mutate(
    gap = as.numeric(timestamp - v_time, units = "secs"),
    vedba = if_else(gap <= 2, vedba, NA)
  )

alb_joined |> 
  as_tibble() |> 
  select(individual_local_identifier, timestamp, v_time, gap, vedba)
```

Here every fix is within a second of its burst, so all matches are kept.
The check matters more when records are sparse. 

These albatross data contain only one GPS fix per individual but five
acceleration bursts, so by joining to our GPS locations we end up discarding
some of the collected acceleration data. If we instead want to associate a
location to each *burst*, an alternative is to interpolate locations to the
burst times with `move2::mt_interpolate()`.

### Continuous data

Some tags record acceleration continuously rather than in bursts. In these cases,
`vedba()` would return a single value for hours of data. Instead, you can split
the bursts into intervals with `split_imu()` and calculate VeDBA for each
interval (see the examples in `?vedba`). Intervals should be much longer than
the window, since each loses about one window's worth of samples at its edges.
Each interval's start time, from `starts()`, can then be used to link it to GPS
fixes with a rolling join, as above.

## Summarizing over time

A single burst describes a few seconds of behavior. For questions about
longer periods (such as detecting when a tag has stopped moving or an
animal may have died), it helps to summarize the per-burst values over time.

The albatross data cover only one hour, so for this step we switch to the
gulls example data, which span almost a day. We calculate VeDBA as before.
These bursts are only 1 second long, which is too short for a running mean,
so we use the mean of each burst (`window = "burst"`) to estimate the static
component. (Note that values from such short bursts should be treated with
caution; we use them here for demonstration purposes.)

```{r gulls}
gul <- gulls()

gul_v <- gul |>
  mutate(
    acceleration = transform_imu(
      as_acc(gul),
      acc_calibration("ornitela", units = "standard_free_fall")
    ),
    vedba = vedba(acceleration, window = "burst")
  ) |>
  filter(!is.na(vedba)) |>
  sf::st_drop_geometry() |>
  as_tibble() |>
  select(individual_local_identifier, timestamp, vedba)
```

We'll use the [slider](https://slider.r-lib.org/) package to calculate the
mean VeDBA over the preceding 5 hours at each burst:

```{r}
library(slider)

gul_v <- gul_v |>
  group_by(individual_local_identifier) |>
  mutate(
    vedba_smooth = slide_index_dbl(
      as.numeric(vedba),
      timestamp,
      mean,
      .before = lubridate::hours(5)
    ),
    vedba_smooth = set_units(vedba_smooth, "standard_free_fall")
  ) |>
  ungroup()
```

A threshold on this rolling mean can then flag periods of prolonged
inactivity. Suitable thresholds depend on the species and tags in question;
here we use 0.05 *g*:

```{r inactive_plot}
threshold <- set_units(0.05, "standard_free_fall")

ggplot(gul_v, aes(x = timestamp)) +
  geom_line(aes(y = vedba), color = "gray") +
  geom_line(aes(y = vedba_smooth), color = "navy") +
  geom_hline(yintercept = threshold, color = "red") +
  labs(x = "", y = "VeDBA")
```

This gull doesn't have a true mortality event. The periods of reduced
activity are likely overnight roosting.

## References

Shepard, E. L. C., Wilson, R. P., Halsey, L. G., Quintana, F., Gómez Laich,
A., Gleiss, A. C., Liebsch, N., Myers, A. E., & Norman, B. (2008).
Derivation of body motion via appropriate smoothing of acceleration data.
*Aquatic Biology*, 4, 235–241. <https://doi.org/10.3354/ab00104>

Wilson, R. P., Börger, L., Holton, M. D., Scantlebury, D. M., Gómez-Laich,
A., Quintana, F., et al. (2020). Estimates for energy expenditure in
free-living animals using acceleration proxies: A reappraisal. *Journal of
Animal Ecology*, 89, 161–172. <https://doi.org/10.1111/1365-2656.13040>
