---
title: "Programming with IMU bursts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with IMU bursts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
Sys.setlocale("LC_TIME", "en_US.utf8")
```

```{css, echo = FALSE}
.html-widget {
  margin-left: auto !important;
  margin-right: auto !important;
  width: 80% !important;
}
```

After extracting IMU data into a vector of bursts, we typically will want
to summarize the individual burst samples into a more interpretable
measure of activity. This vignette shows how to extract simple activity
metrics from `acc` bursts and reattach these metrics to corresponding location
data collected with GPS.

```{r load, message = FALSE}
library(move2imu)
library(move2)
library(dplyr)
library(units)
```

We'll use the move2imu gulls data for this demo:

```{r data}
gul <- gulls()
```

And then extract our acceleration bursts:

```{r}
gul$acceleration <- as_acc(gul)

gul$acceleration[1:10]
```

## Dynamic body acceleration

As a simple example, imagine if we wanted to compute the vectorial dynamic body acceleration (VeDBA) for these bursts, and link each of those VeDBA values to the closest GPS location record.

VeDBA is computed by:

1. subtracting the per-axis mean of a burst (the static, gravity-driven
component) from each sample, leaving the dynamic component, and
2. taking the mean Euclidean norm of the dynamic component across
samples.

First, we need to convert our raw acceleration values into
physical units. (This is covered in depth in the
[calibration vignette](calibration.html).) We will use the standard
transformation for Ornitela tags here (TODO: what tags are used here?)

```{r}
gul$acceleration <- transform_imu(
  gul$acceleration,
  acc_calibration("ornitela", units = "standard_free_fall")
)
```

Then, we extract the VeDBA values:

```{r}
gul_v <- gul |>
  mutate(vedba = vedba(acceleration)) |>
  filter(!is.na(vedba)) |>
  select(vedba)
```

```{r vedba_plot}
library(ggplot2)

ggplot() +
  geom_line(data = gul_v, aes(x = mt_time(gul_v), y = vedba)) +
  labs(x = "", y = "VeDBA") +
  scale_x_datetime(expand = expansion(mult = 0.1))
```

If we're interested in how these changes are expressed in space, we need to
reattach the computed VeDBA to the recorded GPS locations for our gull. However,
the acceleration data is not directly linked with
a particular location. We can see that these records are associated with
`POINT EMPTY` geometries:

```{r}
gul_v$geometry
```

The tag did record locations, though---it's simply that these records are 
stored in the GPS rows of the input move2 object, not the acceleration rows. 
So, we need to relink our VeDBA values with
the corresponding GPS location at the same timestamp. First, we'll drop the
empty geometry column from our VeDBA data and restrict our location data just
to those records with spatial information:

```{r}
# Drop geometry column, which is empty for our acc records:
gul_v <- sf::st_drop_geometry(gul_v)

# Extract location records
gul_gps <- gul[!sf::st_is_empty(gul), ]
```

Then, we can use `dplyr::left_join()` to reattach the two, linking by 
the track ID and the timestamp. (We use the move2 helpers `mt_track_id_column()`
and `mt_time_column()` to ensure the correct metadata columns are used for the
join, as the precise names of these columns can differ across data sources.)

```{r}
# Specify that the track ID columns and timestamp columns should match,
# regardless of the exact column names:
join <- setNames(
  c(mt_track_id_column(gul_v), mt_time_column(gul_v)),
  c(mt_track_id_column(gul_gps), mt_time_column(gul_gps))
)

join

gul_joined <- left_join(gul_gps, gul_v, by = join)
```

Now we have our VeDBA values associated with their corresponding GPS location:

```{r}
library(leaflet)

pal <- colorNumeric("viridis", as.numeric(gul_joined$vedba))

leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolylines(
    data = mt_track_lines(gul_joined),
    color = "gray",
    weight = 0.5
  ) |>
  addCircles(
    data = gul_joined,
    color = ~ pal(as.numeric(vedba)),
    opacity = 1,
    radius = 3
  ) |>
  addLegend(
    title = "VeDBA (g)",
    pal = pal,
    values = as.numeric(gul_joined$vedba)
  )
```

Note that in many cases the timestamps from the acceleration records and GPS
records will differ. In these cases, you won't be able to match by exact
track ID and timestamp. Rather, you'll need to interpolate spatial locations to
the times of the acceleration data (see `move2::mt_interpolate()`) or 
link acceleration records to a nearby timestamp instead. We walk through the
latter option later in this vignette.

## Detecting prolonged inactivity

A single burst's VeDBA tells us how active the animal was during
that brief burst window, but often we're interested in longer periods of
inactivity; for instance, to determine whether a tag has stopped working, if
an animal has died, and so on. In these cases, we want to ignore brief periods
of low acceleration, which likely just indicate a period of rest.

We can accomplish this by applying a time-aware rolling average to our 
computed VeDBA values. This will show us the mean VeDBA for the most
recent time interval.

We'll use the [slider](https://slider.r-lib.org/) package to apply this 
technique, computing the mean VeDBA for the prior 5 hours at each time point.

```{r}
library(slider)

gul_v <- gul_v |>
  group_by(mt_track_id(gul_v)) |>
  mutate(
    vedba_smooth = slide_index_dbl(
      as.numeric(vedba),
      mt_time(gul_v),
      mean,
      .before = lubridate::hours(5)
    ),
    vedba_smooth = set_units(vedba_smooth, "standard_free_fall")
  )
```

```{r}
ggplot(gul_v) +
  geom_line(aes(x = mt_time(gul_v), y = vedba), color = "gray") +
  geom_line(aes(x = mt_time(gul_v), y = vedba_smooth), color = "navy") +
  labs(x = "", y = "VeDBA") +
  scale_x_datetime(expand = expansion(mult = 0.1))
```

We can use this rolling average to identify when the tag has become inactive
or the animal has died. The decision of what temporal and VeDBA threshold to use
to make this determination will depend on the species and tags in question.
For this example, we use a threshold of 0.05*g*. If the tag remains below this
for 5 hours, we consider the tag to be inactive.

Since this is a contrived example, this gull doesn't have a true mortality 
event. The periods of reduced activity are likely a result of overnight roosting.

```{r inactive_plot}
threshold <- set_units(0.05, "standard_free_fall")

ggplot(gul_v) +
  geom_line(aes(x = mt_time(gul_v), y = vedba), color = "gray") +
  geom_line(aes(x = mt_time(gul_v), y = vedba_smooth), color = "black") +
  geom_hline(yintercept = threshold, color = "red") +
  labs(x = "", y = "VeDBA") +
  scale_x_datetime(expand = expansion(mult = 0.1))
```

<!-- ### A cruder summary: per-burst SD -->

<!-- VeDBA is one way to reduce a burst to a single activity -->
<!-- number, but not the only one. A simpler alternative is the -->
<!-- mean per-axis standard deviation: for a stationary tag this -->
<!-- approaches zero on every axis, with no decomposition step -->
<!-- needed. -->

<!-- ```{r} -->
<!-- gul_sd <- gul |>  -->
<!--   filter(!is.na(acceleration)) |>  -->
<!--   mutate( -->
<!--     burst_sd = purrr::map_dbl( -->
<!--       bursts(a), -->
<!--       function(b) { -->
<!--         if (is.null(b)) return(NA) -->
<!--         mean(apply(b, 2, var)) -->
<!--       } -->
<!--     ), -->
<!--     burst_sd_smooth = slide_index_dbl( -->
<!--       burst_sd, -->
<!--       timestamp, -->
<!--       mean, -->
<!--       .before = lubridate::hours(5) -->
<!--     ) -->
<!--   ) -->
<!-- ``` -->

<!-- Fed into the same rolling-window-and-threshold pipeline, -->
<!-- `burst_sd` produces a classifier with very similar behaviour -->
<!-- on this data. The difference is mostly interpretive: VeDBA has -->
<!-- units of acceleration and a long literature behind it, while -->
<!-- per-burst SD is cheaper to compute and easier to explain. The -->
<!-- broader point is that the per-burst summary and the temporal -->
<!-- classifier are independent choices — you can swap one without -->
<!-- touching the other. -->

## Custom acceleration computations

Treating a burst-level calculation as a snapshot of activity at a given
time is reasonable when acceleration data are collected in short bursts.
When sampling takes place over a longer time, it may be necessary to apply
moving window functions across an individual burst.

To illustrate, we'll switch to the `albatrosses()` dataset, which has slightly
longer bursts than the gulls above:

```{r}
alb <- albatrosses()

alb <- alb |>
  mutate(a = as_acc(alb))

burst_dur(alb$a)

n_samples(alb$a)
```

These bursts were collected with e-obs tags. For simplicity, we'll calibrate
using the third-generation e-obs tag manufacturer defaults. See the
[calibration vignette](calibration.html) for a more detailed discussion of
tag calibration.

```{r}
alb <- alb |>
  mutate(a = transform_imu(a, acc_calibration("eobs", 4200)))

alb$a
```

### Within-burst computations

As a first custom computation, we'll compute VeDBA across a sliding
window *within* a single burst, rather than collapsing the whole burst
to one number. This shows us how acceleration varies moment-to-moment
during the burst.

We first extract the burst matrix along with the sampling frequency and start
time of the burst. These are necessary to reconstruct the timestamps of each
sample within the burst.

Then, we use `slider_index_dbl()` to calculate VeDBA for each 1-second
section of the burst.

```{r}
# Sample burst
b <- bursts(alb$a)[[2]]
fq <- freqs(alb$a)[[2]]
t0 <- starts(alb$a)[[2]]

# Get timestamp of each sample within the burst
samp_t <- t0 + (seq_len(nrow(b)) - 1) / as.numeric(fq)

vedba_roll <- slide_index_dbl(
  seq_len(nrow(b)),
  samp_t,
  function(idx) {
    sub <- b[idx, , drop = FALSE]
    sub <- t(t(sub) - set_units(colMeans(sub), units(sub), mode = "standard"))
    mean(sqrt(rowSums(sub^2)))
  },
  .before = lubridate::seconds(1)
)

vedba_roll
```

```{r}
ggplot() +
  geom_line(aes(x = samp_t[2:length(samp_t)], y = vedba_roll[2:length(vedba_roll)])) +
  geom_hline(aes(yintercept = drop_units(vedba(alb$a[[2]]))), linetype = "dashed") +
  labs(x = "Sample", y = "VeDBA")
```

The horizontal line shows the full-burst VeDBA produced by `vedba()`.
As you can see, the dynamic body acceleration varies quite a bit from this
value across the burst.

### Scaling up

We can apply this computation to every burst in the dataset easily. First we
define an explicit function containing the computation from above:

```{r}
rolling_vedba <- function(acc, window = lubridate::seconds(1)) {
  if (is.na(acc)) {
    return(NA_real_)
  }

  b <- bursts(acc)[[1]]
  fq <- freqs(acc)[[1]]
  t0 <- starts(acc)[[1]]

  samp_t <- t0 + (seq_len(nrow(b)) - 1) / as.numeric(fq)

  slide_index_dbl(
    seq_len(nrow(b)),
    samp_t,
    function(idx) {
      sub <- b[idx, , drop = FALSE]
      sub <- t(t(sub) - set_units(colMeans(sub), units(sub), mode = "standard"))
      mean(sqrt(rowSums(sub^2)))
    },
    .before = window
  )
}
```

And then we iterate over the `acc` bursts for all
the albatrosses in the study, applying our `rolling_vedba()` function:

```{r}
alb <- alb |>
  mutate(v = purrr::map(a, rolling_vedba))
```

This leaves us with a vector of values for each burst. We can summarize the
variability in VeDBA within each burst by taking the standard deviation of the
rolling values:

```{r}
alb <- alb |>
  mutate(v_sd = purrr::map_dbl(v, sd))

alb$v_sd
```

### Linking acceleration to GPS by nearest timestamp

Once again, we need to link these values to their associated GPS locations if
we are to answer any questions about how activity varies over space.

First, we split our location records from our acceleration records.

```{r}
# Select GPS records and remove acceleration info (will be reattached)
alb_gps <- alb[!sf::st_is_empty(alb), ] |>
  select(-a, -v_sd)

# Select only acceleration records
alb_acc <- alb |>
  select(a, v_sd) |>
  sf::st_drop_geometry() |>
  filter(!is.na(a))
```

With the gulls data, we were able to join by exact timestamp match because
acceleration records and GPS records were recorded simultaneously.
For the albatross data, the GPS records are typically recorded just after
the start of the acceleration data. This means that a simple `left_join()` 
doesn't manage to match GPS and acceleration records:

```{r}
# Doesn't attach--timestamps don't align in all cases
left_join(alb_gps, alb_acc) |>
  select(a, v_sd)
```

Instead, we need to link the acceleration values to the preceding GPS timestamp.
We can use `dplyr::join_by()` to specify a rolling join using the `closest()`
helper. This matches each GPS record to the most recent acceleration burst that
preceded it (using `timestamp <= timestamp`). We also match on
the animal ID to ensure that records aren't linked across individuals.

```{r}
alb_joined <- left_join(
  alb_gps,
  alb_acc,
  by = join_by(
    individual_local_identifier == individual_local_identifier,
    closest(timestamp <= timestamp)
  ),
  suffix = c("", ".y") # Ensure that move2 metadata columns in alb_gps are not renamed
)
```

Each GPS row now carries the acceleration burst (and our `v_sd`
summary) from the most recent preceding burst, allowing the acceleration
summaries to be explored across space:

```{r}
alb_joined |>
  select(a, v_sd)
```

