---
title: "Programming with acceleration bursts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with acceleration bursts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

```{css, echo = FALSE}
.html-widget {
  margin-left: auto !important;
  margin-right: auto !important;
  width: 80% !important;
}
```

After extracting IMU data into a vector of bursts (see
[Loading IMU data](move2imu.html#loading-imu-data)), we will typically want
to summarize the individual burst samples into a more interpretable
measure of activity. This vignette shows how to extract simple activity
metrics from `acc` bursts and reattach these metrics to corresponding location
data collected with GPS.

```{r load, message = FALSE}
library(move2imu)
library(move2)
library(dplyr)
library(units)
library(ggplot2)
```

First, we'll load the albatross example data and extract the acceleration
bursts:

```{r alb}
alb <- albatrosses()
alb$acceleration <- as_acc(alb)
```

These data contain 12-second bursts recorded at 5 Hz across two axes.

```{r}
alb$acceleration
```

## Computing dynamic body acceleration

Dynamic body acceleration is a commonly used measure of
animal activity and is composed of two acceleration components:

- A *static* component, which stems from gravity or other sources of
constant acceleration (and therefore reflects the animal's posture).
- A *dynamic* component, which stems from the animal's own movements.

The dynamic component is a proxy for activity and, in many species, for energy
expenditure (Wilson et al., 2020). To estimate it, we must isolate and remove
the static acceleration to the best of our ability.
We'll demonstrate how to compute vectorial dynamic body acceleration (VeDBA)
on our albatross data.

First, we calibrate our data to units of *g* to make our DBA results more
interpretable. For this purpose, we suppose that all the birds in the study
carried first-generation e-obs tags set to low sensitivity.

```{r}
alb$acceleration <- transform_imu(
  alb$acceleration,
  acc_calibration(
    "eobs",
    tag_id = 1000,
    sensitivity = "low",
    units = "standard_free_fall"
  )
)
```

### Choosing a smoothing window

To separate the two acceleration components for a burst of data, we compute a
running mean over all the samples within a given window of time. This helps
to separate cyclical patterns (e.g. wingbeats) from stable acceleration signals.

In general, the window length should span several cycles of the primary
movement of interest, but remain short enough to avoid misclassifying
changes in posture as movements. Shepard et al. (2008) suggest at least a
3-second window for species whose stroke cycles are shorter than this.
Albatrosses mostly soar rather than flap, but the repeating movements that do
appear in these data (likely including occasional wingbeats) last about a
second or less, so we will adopt this guidance here.

To compute VeDBA, we simply use the `vedba()` function with the chosen
window.

```{r}
alb$vedba <- vedba(alb$acceleration, window = set_units(3, "s"))

alb$vedba
```

It's also possible to identify an appropriate window by computing VeDBA at
multiple window lengths and finding the point at which the curve flattens:

```{r sensitivity}
alb_acc <- alb$acceleration[!is.na(alb$acceleration)]

# Windows of 3 to 31 samples at 5 Hz
windows <- c(3, 5, 7, 11, 15, 21, 31) / 5

# Median VeDBA at each window size
v <- purrr::map_dbl(
  windows,
  function(w) median(as.numeric(vedba(alb_acc, window = w)))
)

d <- data.frame(window = windows, vedba = v)

ggplot(d, aes(x = window, y = vedba)) +
  geom_line() +
  geom_point() +
  labs(x = "Window (s)", y = "Median VeDBA (g)")
```

VeDBA rises steeply for windows up to about 1.5 seconds and changes little
beyond about 3 seconds, which supports choosing a 3-second window.

### Caveats

Some data are collected in shorter bursts that may not fit a meaningful window.
In these cases, you can pass `window = "burst"`, which estimates the static
component as the mean of each whole burst. However, these DBA results should
be interpreted with caution.

Furthermore, VeDBA is sensitive not only to the window length but also to
other features of the data, like sampling rate and number of axes.

These data were sampled at 5 Hz, which is quite slow relative to
most flapping frequencies. This limits the resolution with which we can
identify flapping flight, leading to underestimated VeDBA. Additionally, these
data were recorded on only 2 axes, while DBA metrics are often calculated on
data with 3. These VeDBA values thus shouldn't be compared to
measurements computed on 3-axis data.

The code here is intended to serve as a demonstration of a workflow for
computing these metrics rather than as a comprehensive analysis.

## Linking to locations

Often, the recording times of acceleration and GPS data don't match exactly.
In the albatross data, for instance, some GPS fixes are recorded up to a second
after the start of an acceleration burst. However, we may be interested in
associating acceleration metrics with nearby GPS locations.

To do so, we'll separate the VeDBA values
calculated above from the GPS fixes by removing rows with missing VeDBA values:

```{r}
alb_v <- alb |>
  filter(!is.na(vedba)) |>
  sf::st_drop_geometry() |>
  as_tibble() |>
  select(individual_local_identifier, timestamp, vedba) |>
  rename(vedba_time = timestamp)

head(alb_v)
```

Then, we'll isolate the GPS location rows by removing rows with empty geometries
and dropping the acceleration-based columns:

```{r}
alb_gps <- alb |>
  filter(!sf::st_is_empty(alb)) |>
  select(-acceleration, -vedba)
```

We can then join our VeDBA values to our GPS fix locations by each animal's
ID and recording timestamp. However, if we
require timestamps to match exactly, only 2 of the 9 fixes would have a match:

```{r}
left_join(
  alb_gps,
  alb_v,
  by = join_by(individual_local_identifier, timestamp == vedba_time)
) |>
  pull(vedba)
```

Instead, we can use a *rolling join* to match all records by a provided
inequality. Here, `closest(timestamp >= vedba_time)` links records by matching
each GPS fix timestamp to the most recent VeDBA record that started at or
before it, within the same individual:

```{r}
alb_joined <- left_join(
  alb_gps,
  alb_v,
  by = join_by(individual_local_identifier, closest(timestamp >= vedba_time))
)
```

A rolling join always takes the closest match, no matter how far away in time
it is. Before using the result, you may want to discard matches where the records
are far from each other in time. For instance, here we compute the time difference
for each match and keep only cases where the difference is at most 2 seconds:

```{r}
alb_joined <- alb_joined |>
  mutate(
    gap = as.numeric(timestamp - vedba_time, units = "secs"),
    vedba = if_else(gap <= 2, vedba, NA)
  )

alb_joined |>
  as_tibble() |>
  select(individual_local_identifier, timestamp, vedba_time, gap, vedba)
```

Here every fix is within a second of its burst, so all matches are kept.
The check matters more when records are sparse.

### Interpolating locations

These albatross data contain only one GPS fix per individual but five
acceleration bursts, so by joining to our GPS locations we end up discarding
some of the collected acceleration data. If we instead want to associate a
location with each *burst*, an alternative is to interpolate locations to the
burst times with `move2::mt_interpolate()`.

### Continuous data

Some tags record acceleration continuously rather than in bursts. In these cases,
`vedba()` would return a single value for hours of data. Instead, you can
separate the dynamic component of the whole recording with `dynamic_acc()`,
then split it into intervals with `split_imu()` and calculate VeDBA for each
interval (see the examples in `?dynamic_acc`). Each interval's start time 
can then be used to link it to GPS fixes with a rolling join, as above.

Separating the dynamic component
before splitting means that only the samples at the two ends of the full
recording are lost when calculating the static acceleration, rather than the
samples at the edges of every interval.

## References

Shepard, E. L. C., Wilson, R. P., Halsey, L. G., Quintana, F., Gómez Laich,
A., Gleiss, A. C., Liebsch, N., Myers, A. E., & Norman, B. (2008).
Derivation of body motion via appropriate smoothing of acceleration data.
*Aquatic Biology*, 4, 235–241. <https://doi.org/10.3354/ab00104>

Wilson, R. P., Börger, L., Holton, M. D., Scantlebury, D. M., Gómez-Laich,
A., Quintana, F., et al. (2020). Estimates for energy expenditure in
free-living animals using acceleration proxies: A reappraisal. *Journal of
Animal Ecology*, 89, 161–172. <https://doi.org/10.1111/1365-2656.13040>
